# Supseded functions

## Smoothing data {#sec-smoothing}


```{r results='asis', echo=FALSE}
function_structure("smooth_cases")
```

The `smooth_cases` function provides procedures to smooth single-case data and eliminate noise. A moving average function (mean- or median-based) replaces each data point by the average of the surrounding data points step-by-step. A *lag* defines the number of measurements before and after the calculation is based on. So a lag-1 will take the average of the proceeding and following value and lag-2 the average of the two proceeding and two following measurements. With a local regression function, each data point is regressed by its surrounding data points. Here, the proportion of measurements surrounding a value is usually defined. So an intensity of 0.2 will take the surrounding 20% of data as the basis for a regression.\
The function returns am scdf with smoothed data points.

```{r smooth_example, fig.width=5, fig.height=5}
## Use the three different smoothing functions and compare the results
berta_mmd <- smooth_cases(Huber2014$Berta)
berta_mmn <- smooth_cases(Huber2014$Berta, FUN = "movingMean")
berta_lre <- smooth_cases(Huber2014$Berta, FUN = "localRegression")
new_study <- c(Huber2014$Berta, berta_mmd, berta_mmn, berta_lre)
names(new_study) <- c("Original", "Moving Median", "Moving Mean", "Local Regression")
plot(new_study, style = "grid2")
```

Here is the syntax for the upcoming `scplot()` function (see @sec-scplot):

```{r fig.width=5, fig.height=5}

scplot(new_study) %>% add_ridge()
```
