
# Exporting *scan* results

```{r results='asis', echo=FALSE}
function_structure("export")
```

The `export` function will make it easier to convert the results of your `scan` analyses into tables and descriptions you can add to your documents and presentations. Basically, `export` takes a `scan` object and converts it to an html-table or latex output.

```{block, type = 'rmdnote'} 
`export` it build on top of the `knitr` and `kableextra` packages. The list provided in the `kable_options` argument is implemented in the `kable` function of `knitr` and the list provided to the `kable_styling_options` is implemented in the `kable_styling` command of the `kableExtra` package. `export` sets some defaults for these functions but you can play around and overwrite them.  
```

`export` works best when used within an rmarkdown file and/or within `RStudio`.  
In `RStudio` 

[xxx to be continued!]

## Single case data files

```{r warning=FALSE}
export(exampleA1B1A2B2_zvt)
```

## Descriptive stats

```{r warning=FALSE}
res <- describeSC(GruenkeWilbert2014)
export(res)
```

## Overlap indices

```{r warning=FALSE}
res <- overlap(exampleA1B1A2B2_zvt, phases = list(c(1,3),c(2,4)))
export(res, flip = TRUE)
```

## Piecewise linear models

```{r warning=FALSE}
res <- plm(exampleA1B1A2B2$Pawel)
export(res)
```

## Hierarchical piecewise regressions

```{r warning=FALSE}
res <- hplm(exampleAB_50, data.l2 = exampleAB_50.l2, lr.test = TRUE, random.slopes = TRUE)
export(res)
```
