# Helper functions

## Defining variables for analysis

In an scdf it is defined which of the variables are the measurement time, phase and study variable. If you have multiple measurement time, phases or dependent variables, then you can switch between them with some helper functions:
`set_dvar` changes the dependent variable, `set_mvar` the measurement-time variable, and `set_pvar` the phase variable. `set_vars` is a redundant function so set several variables at once.

```{r}
#| results: asis
#| echo: false
function_structure("set_dvar")
```

```{r}
#| results: asis
#| echo: false
function_structure("set_mvar")
```

```{r}
#| results: asis
#| echo: false
function_structure("set_pvar")
```

```{r}
#| results: asis
#| echo: false
function_structure("set_vars")
```

Assume we have a single-case study and smooth the dependent variable:

```{r}
scdf <- Huber2014$Berta %>%
  transform(compliance_smooth = local_regression(compliance))
```

Now we change the dependent variable to `compliance_smooth` for an analysis:

```{r}
scdf %>% 
  set_dvar("compliance_smooth") %>%
  smd()
```

If you want to change the variables permanently, replace an scdf with its changed version:

```{r}
scdf <- set_dvar(scdf, "compliance_smooth")
```


