# Setting variables for analysis {#sec-setvars}

In an *scdf*, it is specified which variables represent the measurement time, phase, and dependent variable. If multiple measurement times, phases, or dependent variables are present, they can be switched using helper functions. The function `set_dvar` changes the dependent variable, `set_mvar` modifies the measurement-time variable, and `set_pvar` adjusts the phase variable. The function `set_vars` serves as a redundant option to set multiple variables at once.

```{r}
#| results: asis
#| echo: false
function_structure("set_dvar")
```

```{r}
#| results: asis
#| echo: false
function_structure("set_mvar")
```

```{r}
#| results: asis
#| echo: false
function_structure("set_pvar")
```

```{r}
#| results: asis
#| echo: false
function_structure("set_vars")
```

Assume we have a single-case study and smooth the dependent variable:

```{r}
scdf <- Huber2014$Berta |> 
  transform(compliance_smooth = local_regression(compliance))
```

Now we change the dependent variable to `compliance_smooth` for an analysis:

```{r}
scdf |> 
  set_dvar("compliance_smooth") |> 
  smd()
```

If you want to change the variables permanently, replace an scdf with its changed version:

```{r}
scdf <- set_dvar(scdf, "compliance_smooth")
```


