# Missing values and outliers

## Missing values

There are two kinds of missing values in single-case data series. First, missings that were explicitly recorded as `NA` and assigned to a phase and measurement-time as in the following example:

    scdf(c(5, 3, 4, 6, 8, 7, 9, 7, NA, 6), phase_design = c(A = 4, B = 6))

The second type of missing occurs when there are gaps between measurement-times that are not explicitly coded as in the following example:

    scdf(c(5, 3, 4, 6, 8, 7, 9, 7, 6), phase_design = c(A = 4, B = 5), 
         mt = c(1, 2, 3, 4, 5, 6, 7, 8, 10))

In both cases, missing values pose a threat to the internal validity of overlap indices. Randomization tests are more robust against the first type of missing values but are affected by the second type. Regression approaches are less impacted by both types as they take the interval between measurement-times into account.

```{r}
#| label: fillmissing-example
#| fig-width: 5
#| fig-height: 5
case1 <- scdf(c(3,6,2,4,3,5,2,6,3,2, 6,7,5,8,6,7,4,8,5,6), 
              phase_design = c(A = 10, B = 10), name = "no NA")
case2 <- scdf(c(3,6,2,4,3,5,2,NA,3,2, 6,7,5,8,6,NA,4,8,5,6), 
              phase_design = c(A = 10, B = 10), name = "NAs")
case3 <- fill_missing(case2)
names(case3) <- "interpolated NAs"
ex <- c(case1, case2, case3)
scplot(ex)
overlap(ex)
```

## Outlieranalysis

```{r results='asis', echo=FALSE}
function_structure("outlier")
```

*scan* provides several methods for analyzing outliers. All of them are implemented in the `outliers` function. Available methods are the **standard deviation**, **mean average deviation**, **confidence intervals**, and **Cook's distance**. The criteria argument takes a vector with two information, the first defines the analyzing method ("SD", "MAD", CI", "Cook") and the second the criteria. For "SD" the criteria is the number of standard deviations (**sd**) from the mean of each phase for which a value is not considered to be an outlier. For example, `criteria = c("SD", 2)` would identify every value exceeding two **sd** above or below the mean as an outlier whereas **sd** and mean refer to phase of a value. As this might be misleading particularly for small samples Iglewicz and Hoaglin @iglewicz_how_1993 recommend the use the much more robust median average deviation (**MAD**) instead. The **MAD** is is constructed similar to the **sd** but uses the median instead of the mean. Multiplying the **MAD** by 1.4826 approximates the **sd** in a normal distributed sample. This corrected MAD is applied in the `outlier` function. A deviation of 3.5 times the corrected **MAD** from the median is suggested to be an outlier. To use this criterion set `criteria = c("MAD", 3.5)`. `criteria = c("CI", 0.95)` takes exceeding the 95% confidence interval as the criteria for outliers. The Cook's distance method for calculation outliers can be applied with a strict AB-phase design. in that case, the Cook's distance analyses are based on a piecewise-regression model. Most commonly, Cook's distance exceeding 4/n is used as a criteria. This could be implemented setting `criteria = c("Cook", "4/n")`.

```{r outlier}
outlier(exampleABC_outlier, criteria = c("MAD", 3.5))

# Visualizing outliers with the plot function
res <- outlier(exampleABC_outlier, criteria = c("MAD", 3.5))

scplot(exampleABC_outlier) |>
  add_labels(nudge_y = 15, text = list(size = 0.7)) |>
  add_marks(positions = res) |>
  set_theme("basic") |>
  set_casenames(position = "strip") |>
  set_yaxis(limits = c(20, 120))
```

