# Randomization tests

```{r results='asis', echo=FALSE}
function_structure("rand_test")
```

The `rand_test` function computes a randomization test for single or multiple baseline single-case data. The function is based on an algorithm from the SCRT package (Bulte & Onghena, 2009, 2012), but rewritten and extended for the use in AB designs.

The `statsitics` argument defines the statistic on which the comparison of the phases is based on. The following comparisons are possible:

- "Mean A-B": Uses the difference between the mean of phase A and the mean of phase B. * This is appropriate if a decrease of scores is expected for phase B.
- "Mean B-A": Uses the difference between the mean of phase B and the mean of phase A. This is appropriate if an increase of scores is expected for phase B.
- "Mean |A-B|": Uses the absolute value of the difference between the means of phases A and B.
- "Median A-B": The same as "Mean A-B", but based on the median.
- "Median B-A": The same as "Mean B-A", but based on the median.

*number*  	
Sample size of the randomization distribution. The exactness of the p-value can not exceed 1/number (i.e., number = 100 results in p-values with an exactness of one percent). Default is number = 500. For faster processing use number = 100. For more precise p-values set number = 1000.  

*complete*  	
If TRUE, the distribution is based on a complete permutation of all possible starting combinations. This setting overwrites the number Argument. The default setting is FALSE.

*limit*	 
Minimal number of data points per phase in the sample. The first number refers to the A-phase and the second to the B-phase (e.g., limit = c(5, 3)). If only one number is given, this number is applied to both phases. Default is limit = 5.

*startpoints*  	
Alternative to the limit-parameter, startpoints exactly defines the possible start points of phase B (e.g., startpoints = 4:9 restricts the phase B start points to measurements 4 to 9. startpoints overwrite the limit-parameter.

*exclude.equal*  	
If set to FALSE, which is the default, random distribution values equal to the observed distribution are counted as null-hypothesis conform. That is, they decrease the probability of rejecting the null-hypothesis (increase the p-value). exclude.equal should be set to TRUE if you analyse one single-case design (not a multiple baseline data set) to reach a sufficient power. But be aware, that it increases the chance of an alpha-error.

*graph*  	
If set TRUE, a histogram of the resulting distribution is plotted.

*phases*  	
A vector of two characters or numbers indicating the two phases that should be compared. E.g., phases = c("A","C") or phases = c(2,4) for comparing the second and the fourth phase. Phases could be combined by providing a list with two elements. E.g., phases = list(A = c(1,3), B = c(2,4)) will compare phases 1 and 3 (as A) against 2 and 4 (as B). Default is phases = c("A","B").

```{r rand}
rand_test(exampleAB, graph = TRUE)
```
