# The scan package

## Installing the *scan* package

You can use the `install.packages` function to install *scan*.

`install.packages("scan")` will install the stable version.

The current stable release is version `0.55`. Please look at Section *Software reference* for which version of *scan* has been used for creating this book and make sure you have this version or a newer one installed.

R contains many packages and it would significantly slow down if all packages would be loaded into the computer memory at the beginning of each R session. Therefore, after installing *scan* it needs to be activated at the beginning of each session you use R. Usually a session starts when you start the R program and ends with closing R.

For activating a package you need the `library` function. In this case `library(scan)`. You should get something like

```{r}
#| echo: false
#| results: markup

cat(scan:::.opt$startup_message)
```

indicating that everything went smoothly and *scan* is ready for the job.

## Development version of scan

Alternatively, you can compile the development version of *scan* yourself. This might be necessary if the stable version has some bugs or missing functions which has been fixed.

You may need some computer expertise to get the development version running. It is hosted on gitHub at [\<https://github.com/jazznbass/scan\>](https://github.com/jazznbass/scan){.uri}.

For installation, you can apply the `install_github` function from the `devtools` package (make sure you have installed the `devtools` package before):

`devtools::install_github("jazznbass/scan", dependencies = TRUE)`

When you are running a Windows operating system you will probably have to install Rtools before. Rtools contains additional programs (e.g. compilers) that are needed to compile R source packages.

You can find Rtools here: [\<https://cran.r-project.org/bin/windows/Rtools/\>](https://cran.r-project.org/bin/windows/Rtools/){.uri}

## Reporting issues with *scan* and suggesting enhancements

The *scan* gitHub repository at [\<https://github.com/jazznbass/scan\>](https://github.com/jazznbass/scan){.uri} is the ideal place to report bugs, problems, or ideas for enhancing *scan*. Please use the issue tool (direct link: [\<https://github.com/jazznbass/scan/issues\>](https://github.com/jazznbass/scan/issues){.uri}).

We are very thankful for any feedback, corrections, or whatever helps to improve *scan*!

## Functions overview

The functions of the *scan* package can be divided into the following categories:\
*Manage data*, *analyze*, *manipulate*, *simulate*, and *depict*.

The following tables give an overview of all functions. Furthermore, you can see the current life cycle stage of a function. The life cycle stage categorization is based on the tidyverse package and described in detail here https://lifecycle.r-lib.org/articles/stages.html.

### Management

```{r}
#| echo: false
#| label: tbl-functions-management
#| tbl-cap: Functions for data management
#| results: asis

out <- tribble(
  ~"Function", ~"What it does ...", ~"Lifecycle stage", ~"Chapter",
  #
  "scdf", "Creates a single-case data-frame", "Stable", "@sec-scdf",
  "select_cases", "Selects specific cases of an scdf", "Stable", "@sec-select-cases)",
  "select_phases", "Selects and/or recombines phases", "Stable", "@sec-select-phases)",
  "subset", "Selects specific measurements or variables of an scdf", "stable", "@sec-subset",
  "read_scdf", "Loads external data into an scdf", "Stable", "@sec-read-scdf)",
  "write_scdf", "Writes scdf into an external file", "Stable", "@sec-write-scdf)",
  "convert", "Converts an scdf object into R syntax", "Stable", "@sec-convert",
  "set_var", "(Re)sets dependent, measurement, and phase variable of an scdf", "Stable", "-",
  "scdf_attr", "Gets and sets attributes of an scdf", "Stable", "-",
  "add_l2", "Adds level-two data to an scdf", "Stable", "@sec-add-l2)",
  "as.data.frame/as.data.frame.scdf", "Transforms an scdf into a data frame", "Stable", "-"
)

kable(out,format = "pipe")

#print_table(out)

```

### Depiction

```{r}
#| echo: false
#| label: tbl-functions-depiction
#| tbl-cap: Functions for data depiction/visualisation

out <- tribble(
  ~"Function", ~"What it does ...", ~"Lifecycle stage",
  #
  "plot/plot.scdf", "Creates plots of single cases", "Superseded",
  "scplot", "Add-on package `scplot`. Create advanced ggplot2 plots.", "Experimental",
  "style_plot", "Defines single-case plot graphical styles", "Superseded",
  "export", "Creates html or latex tables from the output of various can functions", "Experimental",
  "print/print.sc", "Prints the results of various scan outputs", "Stable",
  "print/print.scdf", "Prints an scdf", "Stable",
  "summary/summary.scdf", "Summaizes an scdf", "Stable",
  "plot_rand", "Create a distribution plot from a randomization test obejct", "Experimental"
)

print_table(out)

```

### Analysis

```{r}
#| echo: false
#| label: tbl-functions-analysis
#| tbl-cap: Functions for data analysis

out <- tribble(
  ~"Function", ~"What it does ...", ~"Lifecycle stage",
  #
  "autocorr", "Autocorrelations for each phase of each case", "Stable",
  "corrected_tau", "Baseline corrected tau", "Stable",
  "describe", "Descriptive statistics for each phase of each case", "Stable",
  "overlap", "An overview of overlap indeces for each case", "Stable",
  "smd", "Various standardized mean differences between phase A and B", "Stable",
  "rci", "Reliable change index", "Experimental",
  "rand_test", "Randomization test", "Stable",
  "tau_u", "Tau-U for each case and all cases", "Stable",
  "trend", "Trend analyses for each case", "Stable",
  "plm", "Piecewise linear regression model", "Stable",
  "mplm", "Multivariate piecewise linear regression model",  "Experimental",
  "hplm", "Hierarchical piecewise linear regression model",  "Stable",
  "nap", "Non-overlap of all pairs for each case", "Stable",
  "pnd", "Percentage of non overlapping data for each case",  "Stable",
  "pand", "Percentage of all non overlapping data for all cases",  "Stable",
  "pem", "Percantage exceeding the mean for each case",  "Stable",
  "pet", "Percentage exceeding the trend for each case",  "Stable",
  "cdc", "Conservative dual-criterion test",  "Stable",
  "outlier", "Detect outliers for all cases", "Stable"
)

print_table(out)

```

### Manipulation

```{r}
#| echo: false
#| label: tbl-functions-manipulation
#| tbl-cap: Functions for data manipulation

out <- tribble(
  ~"Function", ~"What it does ...", ~"Lifecycle stage",
  #
  "fill_missing", "Interpolate missign values or missing measurement times", "Stable",
  "ranks", "Covert data into ranked data across all cases", "Stable",
  "transform", "Change and create new variabes", "Stable",
  "smooth_cases", "Smoothes time series data", "Stable",
  "truncate_phase", "Deletes measurements of phases", "Stable",
  "standardize", "Standardizes or centers variables across cases", "Stable"
)

print_table(out)

```

### Simulation

```{r}
#| echo: false
#| label: tbl-functions-simulation
#| tbl-cap: Functions for data simulation

out <- tribble(
  ~"Function", ~"What it does ...", ~"Lifecycle stage",
  #
  "design", "Defines a design of one or multiple single-cases", "Stable",
  "power_test", "Calculates power and alpha error of a specific analyzes for a specific single-case design", "Stable",
  "estimate_design", "Extraxt a deisgn template from an existing scdf", "Experimental",
  "random_scdf", "Creats random single-case studies from a single-case design", "Stable",
  "mcscan", "Add-on package `mcscan`. Create Monte-Carlo designs and analyses with `scan`", "(Upcoming not yet functioning)"
)

print_table(out)

```
