
# Piecewise linear regressions

In a piecewise regression analysis (sometimes called segmented regression) a data-set is split at a specific break point and regression parameters (intercept and slopes) are calculated separately for data before and after the break point. This is done because we assume that at the break point a qualitative change happens affecting intercept and slope.
This approach lends itself perfectly to analyze single-case data which are from a statistical point of view time-series data segmented into phases.
A general model for single-case data based on the piecewise regression approach has been suggested by Huitema and McKean @huitema_design_2000. They refer to two phase single-case designs with a pre-intervention phase containing some measurements before the start of the intervention (A-phase) and an intervention phase containing measurements beginning at the intervention’s start and lasting throughout the intervention (B-phase). In this model, four factors predict the outcome at a specific measurement point: The performance at the beginning of the study, a developmental effect leading to a continuous increase throughout all measurements (trend effect), an intervention effect leading to an immediate and constant increase in performance (level effect), and a second intervention effect that evolves continuously with the beginning of the intervention (slope effect).
*scan* provides an implementation based of this piecewise regression approach. Though the original model is extended by several factors:

* multiple phase designs
* additional (control) variables
* autoregression modelling
* logistic, binomial, and poisson distributed dependent variables and error terms
* multivariate analyzes for analyzing the effect of an intervention on more than one outcome variable.

## The basic plm function

The basic function for applying a regression analyzes to a single-case dataset is `plm`. This function analyzes one single-case. In its simplest way, `plm` takes one argument with an *scdf* object and it returns a full piecewise regression analyzes. 

```{r}
plm(exampleAB$Johanna)
```

### Dummy model

The `model` argument is used to code the *dummy variable*. This *dummy variable* is used to compute the slope and level effects of the *phase* variable.  
The *phase* variable is categorical, identifying the phase of each measurement. Typically, categorical variables are implemented by means of dummy variables. In a piecewise regression model two phase effects have to be estimated: a level effect and a slope effect. The level effect is implemented quite straight forward: for each phase beginning with the second phase a new dummy variable is created with values of zero for all measurements except the measurements of the phase in focus where values of one are set.  

```{r echo = FALSE}
#scdf(c(3,6,4,7, 5,3,4,6,3, 7,5,6,4,8), phase.design = c(A = 4, B = 5, C = 5), dummy_level_B = c(0,0,0,0,1,1,1,1,1,0,0,0,0,0), dummy_level_C = c(0,0,0,0,0,0,0,0,0,1,1,1,1,1), name = "example")
#scdf(c(3,6,4,7, 5,3,4,6,3), phase.design = c(A = 4, B = 5), dummy_level_B = c(0,0,0,0,1,1,1,1,1), name = "AB example")
res <- data.frame(phase = c(rep("A",4),rep("B",5)), values = c(3,6,4,7, 5,3,4,6,3), level_B = c(0,0,0,0,1,1,1,1,1))
kable(res)
```

For estimating the *slope effect* of each phase, another kind of dummy variables have to be created. Like the dummy variables for level effects the values are set to zero for all measurements except the ones of the phase in focus. Here, values start to increase with every measurement until the end of the phase.  
Various suggestions have been made regarding the way in which these values increase. The *B&L-B* model starts with a one at the first measurement of the phase and increases with every measurement while the *H-M* model starts with a zero.

```{r echo = FALSE}
#scdf(c(3,6,4,7, 5,3,4,6,3), phase.design = c(A = 4, B = 5), dummy_level = c(0,0,0,0,1,1,1,1,1), "dummy_slop B&L-M" = c(0,0,0,0,1,2,3,4,5), "dummy_slope H-M" = c(0,0,0,0,0,1,2,3,4),name = "AB example", check.names = FALSE )
res <- data.frame(phase = c(rep("A",4),rep("B",5)), values = c(3,6,4,7, 5,3,4,6,3), level = c(0,0,0,0,1,1,1,1,1), "slope B&L-M" = c(0,0,0,0,1,2,3,4,5), "slope H-M" = c(0,0,0,0,0,1,2,3,4),check.names = FALSE )
kable(res)
```

With single-case studies with more than two phases it gets a bit more complicated. Applying the a fore described models to three phases would result in a comparison of each phase to the first phase (usually the A Phase). That is, regression weights and significance tests will depict differences of each phase to the values of phase A. This might be OK depending on what you are interested in. But in a lot of cases we are more interested in analyzing the effects of a phase compared to the previous one.  
This is achieved applying the *JW* dummy model. In this model, the dummy variable for the level effect is set to zero for all phases preceding the phase in focus and set to one for all remaining measurements. Similar, the dummy variable for the slope effect is set to zero for all phases preceding the one in focus and starts with one for the first measurement of the target phase and increases until the last measurement of the case.  

```{r echo = FALSE}
#scdf(c(3,6,4,7, 5,3,4,6,3, 7,5,6,4,8), phase.design = c(A = 4, B = 5, C = 5), level_B = c(0,0,0,0,1,1,1,1,1,1,1,1,1,1), level_C = c(0,0,0,0, 0,0,0,0,0, 1,1,1,1,1), slope_B = c(0,0,0,0,1,2,3,4,5,6,7,8,9,10), slope_C = c(0,0,0,0,0,0,0,0,0,1,2,3,4,5), name = "ABC example")
res <- data.frame(phase = c(rep("A",4),rep("B",5),rep("B",5)), values = c(3,6,4,7, 5,3,4,6,3, 7,5,6,4,8), level_B = c(0,0,0,0,1,1,1,1,1,1,1,1,1,1), level_C = c(0,0,0,0, 0,0,0,0,0, 1,1,1,1,1), slope_B = c(0,0,0,0,1,2,3,4,5,6,7,8,9,10), slope_C = c(0,0,0,0,0,0,0,0,0,1,2,3,4,5))
kable(res)
```


## Different modells

```{r}
plm(exampleAB$Karolina)
```


The piecewise regression reveals a significant level effect and two non significant effects for trend and slope. In a further analyses we would like to put the slope effect out of the equation. There are several ways to do this. The easiest way is the to unset the `slope` argument: `slope = FALSE`.

```{r}
plm(exampleAB$Karolina, slope = FALSE)
```

In the resulting estimations the trend and level effects are now significant. The model estimated a trend effect of 0.96 points per measurement time and a level effect of 14.3 points. That is, with the beginning of the intervention (the B-phase) the score increases by 14.3 points.

## Modelling autoregression

```{r}
autocorrSC(Grosche2011)
```



## Adding additional predictors

In more complex analyses additional predictors can be included in the piecewise regression model. 

To do this, we have to change the regression formula 'manually' by applying the `update` argument. The `update` argument allows to change the underlying regression formula. To add a new variable named for example `newVar`, set `update = .~. + newVar`. The `.~.` part takes the internally build formula and `+ newVar` adds a variable named `newVar` to the equation.



```{r}
exampleAB_add
plm(exampleAB_add, update = .~. + cigarrets)
```

The formula has two parts divided by a tilde. Left of the tilde is the variable to be predicted and right of it the predictors. A `1` indicates the intercept, the variable `mt` estimates the trend effect, `phaseB` the level effect of the B-phase and the variable `interB` the slope effect of the B-phase. If `formula` is not explicitly defined it is set to `formula = values ~ 1 + mt + phaseB + interB` (assuming an AB-design) to estimate the full piecewise regression model. 


